---
title: "Cluster-aware cross-validation for repeated-measures predictors with binary outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cluster-aware cross-validation for repeated-measures predictors with binary outcomes}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r load}
library(cawCV)
```

## The problem

Clinical prediction models built from repeated-measures data carry an
information-leakage problem that ordinary cross-validation does not address.
When several measurements per subject are present in the data, and ordinary
k-fold cross-validation partitions at the observation level, records from the
same subject are routinely split between the training and the test set. Any
within-subject signal that connects those records, whether the signal arises
from a stable baseline characteristic, a subject-specific tendency that shifts
the predictor up or down, or temporal autocorrelation within a subject's
record, can be exploited by the learner to recover information about the
held-out observations from the training observations of the same subject. The
performance estimate that results is optimistically biased relative to the
performance the model will achieve on subjects who were not represented during
training.

Cluster-aware cross-validation removes this leakage at the source. Each
subject's observations are kept together: either all in a training fold or
all in a test fold. Reported performance then reflects what the model can be
expected to achieve on subjects who were not seen during model development,
which is the question clinical prediction is ordinarily trying to answer. The
methodology underlying `cawCV` is treated in detail in Hagan (2026), which
characterizes the optimism of naive cross-validation and quantifies the
mechanisms involved.

This vignette illustrates the practical workflow the package supports:
diagnose within-subject dependence in the predictors, choose a partitioning
strategy consistent with the diagnostics, and compare cross-validated
performance under cluster-aware and naive strategies on the same data.

## A synthetic example

A small synthetic dataset is generated for illustration. The setting matches
the common neonatal and pediatric scenario in which each subject contributes
several measurements and the outcome is a single subject-level binary event.

```{r data}
set.seed(7777)
n_subjects   <- 100
obs_per_subj <- 10
n_total      <- n_subjects * obs_per_subj
subject_id   <- rep(seq_len(n_subjects), each = obs_per_subj)

# Three predictors, each carrying a different kind of within-subject structure.
# x_baseline is constant within subject, that is, a stable subject attribute.
# x_random_intercept varies within subject around a per-subject mean.
# x_ar1 varies within subject as a stationary AR(1) process with phi = 0.9,
# anchored at 0 for each subject so that the within-subject autocorrelation
# is the dominant feature of the series.
x_baseline_per_subject  <- rnorm(n_subjects, mean = 0, sd = 1)
x_intercept_per_subject <- rnorm(n_subjects, mean = 0, sd = 1)

x_baseline         <- x_baseline_per_subject[subject_id]
x_random_intercept <- x_intercept_per_subject[subject_id] +
                      rnorm(n_total, mean = 0, sd = 0.7)

phi      <- 0.9
innov_sd <- sqrt(1 - phi^2)
x_ar1 <- unlist(lapply(seq_len(n_subjects), function(i) {
  series    <- numeric(obs_per_subj)
  series[1] <- rnorm(1, mean = 0, sd = innov_sd)
  for (t in 2:obs_per_subj) {
    series[t] <- phi * series[t - 1] + rnorm(1, mean = 0, sd = innov_sd)
  }
  series
}))

# Outcome is driven by the within-subject PEAK of x_ar1, not by its
# subject-level mean. This is the relevant pattern for clinical prediction
# from autocorrelated repeated measurements, in which an event is triggered
# by a sustained or extreme excursion rather than by an average level. The
# observation-level predictor x_ar1 therefore carries direct prognostic
# information that varies within subject through time.
x_ar1_peak <- tapply(x_ar1, subject_id, max)

linpred  <- -1.5 + 0.6 * x_baseline_per_subject +
                   0.4 * x_intercept_per_subject +
                   1.5 * x_ar1_peak
prob     <- 1 / (1 + exp(-linpred))
y_subj   <- rbinom(n_subjects, size = 1, prob = prob)
y        <- y_subj[subject_id]

dat <- data.frame(
  subject            = subject_id,
  x_baseline         = x_baseline,
  x_random_intercept = x_random_intercept,
  x_ar1              = x_ar1,
  y                  = y
)
head(dat, 4)
table(outcome = y_subj)
```

The dataset contains `r nrow(dat)` observations on `r n_subjects` subjects,
with each subject contributing `r obs_per_subj` measurements and one
outcome.

## Diagnose before you partition

`predictor_icc()` reports the intraclass correlation of each predictor across
subjects: the share of total variance that lies between subjects rather than
within them. `predictor_lag1()` reports the within-subject lag-1
autocorrelation summarised across subjects. The two diagnostics target the
two main routes by which observation-level cross-validation leaks
information.

```{r diagnostics}
predictor_icc(dat,
              predictors = c("x_baseline", "x_random_intercept", "x_ar1"),
              cluster    = dat$subject)

predictor_lag1(dat,
               predictors = c("x_baseline", "x_random_intercept", "x_ar1"),
               cluster    = dat$subject)
```

The diagnostic profile is as constructed. `x_baseline` is constant within
subject, so its intraclass correlation is at its ceiling and a within-subject
lag-1 autocorrelation is not defined. `x_random_intercept` carries a
substantial intraclass correlation, reflecting the per-subject mean shift,
but its within-subject lag-1 autocorrelation is near zero. `x_ar1` carries a
moderate intraclass correlation and a clearly elevated within-subject lag-1
autocorrelation, the signature of an autocorrelated series. In all three
cases, within-subject dependence is present, and the dependence would be
exploited under naive observation-level cross-validation.

## Compare partitioning strategies

The two strategies compared here are naive observation-level k-fold
partitioning, which ignores subject identity, and subject-level k-fold
partitioning, which keeps each subject's observations together. Both
partitions use five folds and the same seed so the comparison reflects the
strategy and not the random draw.

```{r partitions}
partition_naive <- partition_naive_kfold(dat$subject, k = 5, seed = 1)
partition_subj  <- partition_subject_kfold(dat$subject, k = 5, seed = 1)

partition_naive
partition_subj
```

For each strategy, a logistic regression of the binary outcome on the three
predictors is fit on the training fold and the area under the receiver
operating characteristic curve is computed on the held-out fold. The mean
across folds is the cross-validated estimate of discrimination.

```{r auroc-helper}
# A compact base-R AUROC computation, equivalent to the Mann-Whitney U
# statistic divided by the count of positive-negative pairs. No external
# package is required.
auroc <- function(prob, outcome) {
  pos <- prob[outcome == 1]
  neg <- prob[outcome == 0]
  if (length(pos) == 0 || length(neg) == 0) return(NA_real_)
  pairs <- outer(pos, neg, FUN = function(a, b) {
    (a > b) + 0.5 * (a == b)
  })
  mean(pairs)
}

cv_auroc <- function(partition, data) {
  out <- numeric(n_folds(partition))
  for (f in seq_len(n_folds(partition))) {
    idx <- get_fold(partition, f)
    fit <- glm(y ~ x_baseline + x_random_intercept + x_ar1,
               data   = data[idx$train, ],
               family = binomial())
    p   <- predict(fit, newdata = data[idx$test, ], type = "response")
    out[f] <- auroc(p, data$y[idx$test])
  }
  out
}
```

```{r run-cv}
naive_aurocs <- cv_auroc(partition_naive, dat)
subj_aurocs  <- cv_auroc(partition_subj,  dat)

data.frame(
  strategy       = c("naive observation-level k-fold",
                     "subject-level k-fold"),
  mean_auroc     = c(mean(naive_aurocs), mean(subj_aurocs)),
  sd_across_folds = c(sd(naive_aurocs),  sd(subj_aurocs))
)
```

The naive observation-level strategy returns the higher mean cross-validated
area under the curve on this dataset. The difference is the optimism
introduced by the leakage of within-subject information into the held-out
folds. The subject-level estimate reflects performance on subjects who were
not seen during model development, which is the question clinical prediction
is ordinarily trying to answer. A formal characterization of this optimism,
including its dependence on intraclass correlation, autocorrelation, and the
number of records per subject, is given in Hagan (2026).

A third partitioning strategy, leave-one-cluster-out, is available through
`partition_loco()` for settings in which the number of subjects is small
enough to make an exhaustive cluster-aware comparison practical. Future
package releases will add further pre-validation diagnostics and partitioning
strategies as the underlying methodological work is published.

## Reference

Hagan JL. Quantifying the Optimism of Naive Cross-Validation for Binary
Outcome Prediction with Repeated-Measures Predictors: A Simulation Study and
Clinical Illustration. medRxiv. 2026.
https://doi.org/10.64898/2026.05.27.26354222
